Constraining the Interiors of Asteroids Through Close Encounters

### 2022-05-17 paper/appendices.tex
Corrected threshold line in cadence n cadence appendix

In figure \ref{fig:cad-contour}, we display contour plots of posterior uncertainty $\sigma(K_{\ell m})$ of the fit parameters as a function of both cadence $\Delta t$ and $P_\omega$ (left panel) or the relative orbit speed $t_\text{spin} / t_\text{orbit}$ (right panel). Both panels show the same sudden increase in posterior uncertainty we named $T_\text{cad}$, now visible as a function of frequency and relative orbit speed. In all cases, the value of $\gamma_0$ was set so that all data points achieve the same value of $\gamma$ at perigee.

\begin{figure*}
  \centering
  \includegraphics[width=0.48\textwidth]{figs/cad-period.pdf}\hfill
  \includegraphics[width=0.48\textwidth]{figs/cad-speed.pdf}
  \caption{Contour plots showing posterior uncertainties as a function of cadence $\Delta t$ and dynamical time scales for the encounter: rotational period $P_\omega$ (\textit{left}) and the relative speed of the orbit (\textit{right}; see text for a definition). The reference values of $P_\omega=9$ hr and $t_\text{spin}/t_\text{orbit}=1$ are shown as dotted lines. The red line represents the uncertainty threshold. Cadence cutoff depends strongly on both $P_\text{omega}$ and $t_\text{spin}/t_\text{orbit}=1$.}
  \label{fig:cad-contour}
\end{figure*}

>>>>

In figure \ref{fig:cad-contour}, we display contour plots of posterior uncertainty $\sigma(K_{\ell m})$ of the fit parameters as a function of both cadence $\Delta t$ and $P_\omega$ (left panel) or the relative orbit speed $t_\text{spin} / t_\text{orbit}$ (right panel). Both panels show the same sudden increase in posterior uncertainty we named $T_\text{cad}$, located around the region where $\sigma_\rho / \rho = 100\%$ and now visible as a function of frequency and relative orbit speed. In all cases, the value of $\gamma_0$ was set so that all data points achieve the same value of $\gamma$ at perigee.

\begin{figure*}
  \centering
  \includegraphics[width=0.48\textwidth]{figs/cad-period.pdf}\hfill
  \includegraphics[width=0.48\textwidth]{figs/cad-speed.pdf}
  \caption{Contour plots showing posterior uncertainties as a function of cadence $\Delta t$ and dynamical time scales for the encounter: rotational period $P_\omega$ (\textit{left}) and the relative speed of the orbit (\textit{right}; see text for a definition). The reference values of $P_\omega=9$ hr and $t_\text{spin}/t_\text{orbit}=1$ are shown as dotted lines. The solid (dotted) red line represents the $\sigma_\rho / \rho = 100\%$ (20\%) threshold. Cadence cut-off depends strongly on both $P_\text{omega}$ and $t_\text{spin}/t_\text{orbit}=1$.}
  \label{fig:cad-contour}
\end{figure*}

### 2022-05-17 paper/main.tex
Corrected threshold line in cadence n cadence appendix

\jtd{It will be useful to have a rule-of-thumb ``posterior uncertainty threshold,'' above which we say that posterior uncertainty $\sigma(K_{\ell m})$ is too large to yield meaningful estimates of density distribution. We define this posterior uncertainty threshold as the point at which the median of the ratio between the the local density uncertainty and local density  ($\rho_\sigma / \rho$) is equal to a fraction $f$. For this analysis, we investigate both $f=100\%$ (corresponding to an essentially unconstrained density distribution) and $f=20\%$.}

\jtd{We use this threshold because it has clear physical meaning, unlike the density moments. But the value of $\rho_\sigma / \rho$ depends on the model used. In this analysis, we use the likelihood model, but choosing a different model with constraints added would reduce density uncertainty, and lead to less restrictive thresholds.}

@@

Figure \ref{fig:scan-vex} demonstrates that posterior uncertainty follows a slight trend for uncertainty to increase with $v_\infty$. This is likely due to the fact that larger $v_\infty$ leads to a faster and flatter orbit with less time spent close to the planet, where tidal torque is strongest. There are also comparatively large oscillations in the uncertainty, due to the orientation of the asteroid at perigee varying. (The asteroid is always simulated to start at the same orientation, but increasing $v_\infty$ decreases the time to perigee so that the asteroid enters this region of high torque at different orientations depending on $v_\infty$.) This explains why the oscillations have the same period for all parameters. Note that these oscillations are sometimes large enough to raise $\sigma(K_{3m}) > 0.01$, shown by the red vertical line.

The figure shows much stronger dependence of parameter uncertainty on perigee distance, as expected by the factor of $(a_\mathcal{A}/D)^{\ell'}$ present in equation \ref{eqn:tidal-torque} and mentioned in section \ref{sec:tidal-torque}. For $r_p \approx 6$ or 7 Earth radii, $\sigma(K_{30})$ reaches the 0.01 threshold. At slightly higher $r_p$, $K_{31}$ and $K_{32}$ become similarly unresolved, and at $r_p \approx 20$ Earth radii, $K_{33}$, the last second-order component, becomes unresolved. Even before this, at $r_p \approx 10$ Earth radii, the most uncertain parameter $K_{30}$ fills the prior distribution with uncertainty ranging from -1 to 1, visible by the sudden cut-off in uncertainty increase and the discontinuity of the $\sigma(K_{\ell m})$ curve there. The exact location of these limits will change if different encounter properties are used.

@@

The very strong dependence of $\sigma(K_{\ell m})$ on $r_p$ makes this analysis most useful to extract second-order moments on close encounters (5-20 Earth radii). Fortunately, in the case of Earth, these encounters are also likely to have the best associated observational uncertainty when above the horizon due to their proximity. The first-order moments can still be extracted at much larger perigee distances in our model.

@@

For the case in which $\sigma_\rho$ is held fixed and $\sigma_\theta$ is varied, the figure shows that the dependence of $\sigma(K_{\ell m})$ on $\sigma_\theta$ is linear. The $K_{30}$ parameter reaches the $\sigma(K_{30})\approx 0.01$ threshold at $\sigma_\theta = 0.032\ \text{radians} \approx 2^\circ$, whereas the other parameters reach the threshold between $8^\circ$ and $45^\circ$. The $K_{2m}$ uncertainties remain low, even for $\sigma_\theta \approx 1$, which is large enough that the spin pole has non-negligible probability to be observed in any direction.

If $\sigma_\rho$ is allowed to vary, then $K_{30}$ reaches the threshold at $\sigma_\rho \approx \num{2.8e-7}$, and the other parameters reach the threshold at $\sigma_\rho \approx 1.2-\num{8.6e-6}$. These correspond to uncertainty in period of roughly milliseconds to seconds. At large period uncertainty, the PPDs begin to fill the prior and the posterior uncertainty $\sigma(K_{\ell m}$) is no longer linear with $\sigma_\rho$ (this is especially visible in the $K_{30}$) case. Otherwise, posterior uncertainty is proportional to $\sigma_\rho$.

@@

In figure \ref{fig:scan-space-sigma}, we show the 1$\sigma$ posterior uncertainties as a function of $K_{20}$ and $K_{22}$, or alternatively $a/c$ and $b/c$. We use axis ratios rather than the values of $a$, $b$, and $c$ because axis ratios are independent of $a_\mathcal{A}$. \jtd{The large $|K_{22}|$ sides of the $K_{2m}$-space plots correspond to asymmetric ellipsoids, as do the points far from the $a/c=b/c$ diagonal of the axis-ratio-space plots. The point corresponding to a sphere, which experiences no tidal torque and no tumbling, is at $K_{2m}=0$ and $a/c=b/c=1$.}

@@

\jtd{The fact that $K_{\ell m}$ are correlated indicates that, in order to correctly propagate density moment uncertainties to a density distribution, a full PPD is necessary. If only 1$\sigma$ uncertainty intervals for each parameter were used, density distribution uncertainty would be overestimated. This full PPD is produced by our fit process, but only the 1$\sigma$ uncertainties are displayed in this paper in most cases.}

@@

On the other hand, the posterior uncertainty of $K_{3m}$ is much more strongly dependent on asteroid length $a_\mathcal{A}$. Figure \ref{fig:scan-am} displays posterior uncertainty $\sigma(K_{\ell m})$ as a function of $a_\mathcal{A}$. As was mention in section \ref{sec:tidal-torque}, the $K_{2m}$ parameters are insensitive to $a_\mathcal{A}$ since the $a_\mathcal{A}^2$ term in $\bm \tau$ (equation \ref{eqn:tidal-torque}) cancels the $a_\mathcal{A}^2$ in the moment of inertia (equation \ref{eqn:moi}). The $K_{3m}$ uncertainty is strongly dependent on $a_\mathcal{A}$ for the same reason that it is dependent on $r_p$: the $(a_\mathcal{A}/D)^{\ell'}$ dependence of equation \ref{eqn:tidal-torque}. At $a_\mathcal{A} \lesssim 700$ m, $K_{30}$ becomes unresolved. The other parameters become unresolved for $a_\mathcal{A} \lesssim 300$ m, with the cut-off for $K_{33}$ being less than 100 meters. This $a_\mathcal{A}$ cut-off is quite sharp; figure \ref{fig:scan-am} shows steep decrease in $\sigma(a_\mathcal{A})$ for low $a_\mathcal{A}$, with an approximate functional form of $\sigma(a_\mathcal{A}) \sim e^{(a_\mathcal{A}/C)^p}$ for a scaling constant $C$ and some power $p < 0.1$. This sharpness indicates that the cut-off value of $a_\mathcal{A}$ is unlikely to change significantly if other parameters of the flyby are altered in a way that slightly increases or decreases $\sigma(K_{\ell m})$. It therefore appears that extracting precise $K_{30}$ from asteroids with $a_\mathcal{A} \lesssim 700$ m is unlikely with this analysis, and for $a_\mathcal{A} \lesssim 50$ m, all $K_{3m}$ will likely be unresolved.

@@

\begin{figure*}
  \centering
  \includegraphics[width=0.8\textwidth]{figs/spin-pole.png}
  \caption{$1\sigma$ uncertainties for the first-order parameters (\textit{top}) and second-order (\textit{bottom}) as a function of the initial direction of spin in the inertial frame. All maps are made in the Mollweide projection. The orange star indicates the reference spin pole. The red contours enclose regions where $\sigma(K_{3m}) \geq 0.01$. Posterior uncertainty depends similarly on initial spin pole direction for all parameters}
  \label{fig:scan-spin}
\end{figure*}

\begin{figure*}
  \centering
  \includegraphics[width=0.3\textwidth]{figs/spin-pole-avg-mollweide.png}
  \includegraphics[width=0.3\textwidth]{figs/spin-pole-avg-polar.png}
  \caption{The weighted average of the uncertainties shown in figure \ref{fig:scan-spin}, in Mollweide (\textit{left}) and polar (\textit{right}) projections in the inertial frame. See text for a description of how the average was computed. Black dots indicate the Fibonacci-sphere-distributed locations of sample spin poles, and the orange star indicates the reference spin pole. The polar projection is centred at the south pole, or $-\unit Z$. Posterior uncertainty is large near $\pm \unit Z$ and $\pm \unit Y$, but roughly constant elsewhere.}
  \label{fig:scan-spin-avg}
\end{figure*}

@@

The most severe effect of period on $\sigma(K_{\ell m})$ is in the low-period regime ($P_\omega \lesssim 5$ hr), but in this case, the most strongly affected parameters are $K_{2m}$, which are generally known better than $K_{3m}$. The effect on the imprecise parameters $K_{3m}$ is small, except for $K_{30}$. It therefore seems as though small-period asteroids are still candidates for observation, although high-period asteroids yield smaller uncertainty.

@@

We assessed the posterior uncertainty generated by the fit process by adjusting various properties of an asteroid on an Earth encounter. The main conclusions are listed below.

\begin{enumerate}
  \item Highly precise rotational period data is required to produce precise second-order moments.
  \item Large $a_\mathcal{A}$ is required for $K_{3m}$ resolution, though the size of $a_\mathcal{A}$ has no effect on $K_{2m}$. The threshold on $a_\mathcal{A}$ for the reference asteroid was $\sim$700 m for $K_{30}$ and $\sim$50 m for $K_{33}$.
  \item A close encounter is vital for accurate second-order moment extraction, though some of the second-order moments can still be required for more distant encounters.
  \item This model will yield inflated uncertainties if the asteroid is rotationally symmetric around its rotating spin axis. In this case, a different parametrization should be used. 
  \item If the asteroid rotational period is too small, or if the initial angular velocity points in an inconvenient direction, the density moments will be poorly constrained.
  \item The cadence of observation should generally be shorter than about 30 minutes for the reference asteroid case. Gaps of data can appear, even at perigee, but not for more than about an hour.
\end{enumerate}

If some of these conditions are not met, the first-order $K_{2m}$ moments may still be resolvable.

>>>>

It will be useful to have a rule-of-thumb ``posterior uncertainty threshold,'' above which we say that posterior uncertainty $\sigma(K_{\ell m})$ is too large to yield meaningful estimates of density distribution. We define this posterior uncertainty threshold as the point at which the median of the ratio between the the local density uncertainty and local density  ($\rho_\sigma / \rho$) is equal to a fraction $f$. For this analysis, we investigate both $f=100\%$ (corresponding to an essentially unconstrained density distribution) and $f=20\%$.

We use this threshold because it has clear physical meaning, unlike the density moments. But the value of $\rho_\sigma / \rho$ depends on the model used, unlike $\sigma(K_{\ell m})$. In this analysis, we use the likelihood model, but choosing a different model with constraints added would reduce density uncertainty, and lead to less restrictive thresholds.

@@

Figure \ref{fig:scan-vex} demonstrates that posterior uncertainty follows a slight trend for uncertainty to increase with $v_\infty$. This is likely due to the fact that larger $v_\infty$ leads to a faster and flatter orbit with less time spent close to the planet, where tidal torque is strongest. There are also comparatively large oscillations in the uncertainty, due to the orientation of the asteroid at perigee varying. (The asteroid is always simulated to start at the same orientation, but increasing $v_\infty$ decreases the time to perigee so that the asteroid enters this region of high torque at different orientations depending on $v_\infty$.) Hence, the oscillations have the same period for all parameters. The dependence of posterior uncertainty on $v_\infty$ is weak enough so that $\sigma_\rho/\rho$ never reaches 100\% for the $v_\infty$ investigated.

The figure shows much stronger dependence of parameter uncertainty on perigee distance, as expected by the factor of $(a_\mathcal{A}/D)^{\ell'}$ present in equation \ref{eqn:tidal-torque} and mentioned in section \ref{sec:tidal-torque}. For $r_p =7.9$ Earth radii, $\sigma_\rho/\rho$ reaches the 100\% threshold. The 20\% threshold is reached at lower 4.7 Earth radii. Much of this uncertainty is dominated by uncertainty in $K_{30}$, which is the parameter least constrained by tidal torque analysis. At $r_p \approx 10$ Earth radii, $K_{30}$ fills the prior distribution with uncertainty ranging from -1 to 1, visible by the sudden cut-off in uncertainty increase and the discontinuity of the $\sigma(K_{\ell m})$ curve there.

@@

The very strong dependence of $\sigma(K_{\ell m})$ on $r_p$ makes this analysis most useful to extract second-order moments on close encounters. Fortunately, in the case of Earth, these encounters are also likely to have the best associated observational uncertainty when above the horizon due to their proximity. The first-order moments can still be extracted at much larger perigee distances in our model.

@@

For the case in which $\sigma_\rho$ is held fixed and $\sigma_\theta$ is varied, the figure shows that the dependence of $\sigma(K_{\ell m})$ on $\sigma_\theta$ is linear. The uncertainty threshold is reached at $\sigma_\theta = 2.8^\circ$ (0.5$^\circ$) for $\sigma_\rho / \rho=$100\% (20\%). The $K_{2m}$ uncertainties remain low, even for $\sigma_\theta \approx 1$, which is large enough that the spin pole has non-negligible probability to be observed in any direction.

If $\sigma_\rho$ is allowed to vary instead, then $K_{30}$ reaches the 100\% and 20\% thresholds at $\sigma_\rho$ which correspond to period uncertainty of several milliseconds. At large period uncertainty, the PPDs begin to fill the prior and the posterior uncertainty $\sigma(K_{\ell m}$) is no longer linear with $\sigma_\rho$ (this is especially visible in the $K_{30}$) case. Otherwise, posterior uncertainty is proportional to $\sigma_\rho$.

@@

In figure \ref{fig:scan-space-sigma}, we show the 1$\sigma$ posterior uncertainties as a function of $K_{20}$ and $K_{22}$, or alternatively $a/c$ and $b/c$. We use axis ratios rather than the values of $a$, $b$, and $c$ because axis ratios are independent of $a_\mathcal{A}$. The large $|K_{22}|$ sides of the $K_{2m}$-space plots correspond to asymmetric ellipsoids, as do the points far from the $a/c=b/c$ diagonal of the axis-ratio-space plots. The point corresponding to a sphere, which experiences no tidal torque and no tumbling, is at $K_{2m}=0$ and $a/c=b/c=1$.

@@

The fact that $K_{\ell m}$ are correlated indicates that, in order to correctly propagate density moment uncertainties to a density distribution, a full PPD is necessary. If only 1$\sigma$ uncertainty intervals for each parameter were used, density distribution uncertainty would be overestimated. This full PPD is produced by our fit process, but only the 1$\sigma$ uncertainties are displayed in this paper in most cases.

@@

On the other hand, the posterior uncertainty of $K_{3m}$ is much more strongly dependent on asteroid length $a_\mathcal{A}$. Figure \ref{fig:scan-am} displays posterior uncertainty $\sigma(K_{\ell m})$ as a function of $a_\mathcal{A}$. As was mention in section \ref{sec:tidal-torque}, the $K_{2m}$ parameters are insensitive to $a_\mathcal{A}$ since the $a_\mathcal{A}^2$ term in $\bm \tau$ (equation \ref{eqn:tidal-torque}) cancels the $a_\mathcal{A}^2$ in the moment of inertia (equation \ref{eqn:moi}). The $K_{3m}$ uncertainty is strongly dependent on $a_\mathcal{A}$ for the same reason that it is dependent on $r_p$: the $(a_\mathcal{A}/D)^{\ell'}$ dependence of equation \ref{eqn:tidal-torque}. At $a_\mathcal{A} \leq 1100$ m, $\sigma_\rho/\rho \approx 100\%$. However, the 20\% cut-off is much lower, at $a_\mathcal{A} \leq 180$. Figure \ref{fig:scan-am} demonstrates why these cut-offs are so distant; for large $a_\mathcal{A}$, uncertainty decreases slowly with length. Only for small lengths is the uncertainty very large. This behaviour indicates that the uncertainty threshold is unlikely to fall much below 180 m even if other properties of the encounter are adjusted.

@@

\begin{figure*}
  \centering
  \includegraphics[width=0.8\textwidth]{figs/spin-pole.png}
  \caption{$1\sigma$ uncertainties for the first-order parameters (\textit{top}) and second-order (\textit{bottom}) as a function of the initial direction of spin in the inertial frame. All maps are made in the Mollweide projection. The orange star indicates the reference spin pole. The red contours enclose regions the 20\% (dotted) and 100\% (solid) $\sigma_\rho / \rho$ cut-off. Posterior uncertainty depends similarly on initial spin pole direction for all parameters}
  \label{fig:scan-spin}
\end{figure*}

\begin{figure*}
  \centering
  \includegraphics[width=0.3\textwidth]{figs/spin-pole-avg-mollweide.png}
  \includegraphics[width=0.3\textwidth]{figs/spin-pole-avg-polar.png}
  \caption{The weighted average of the uncertainties shown in figure \ref{fig:scan-spin}, in Mollweide (\textit{left}) and polar (\textit{right}) projections in the inertial frame. See text for a description of how the average was computed. Black dots indicate the Fibonacci-sphere-distributed locations of sample spin poles, and the orange star indicates the reference spin pole. The polar projection is centred at the north pole, or $\unit Z$. Posterior uncertainty is large near $\pm \unit Z$ and $\pm \unit Y$, but roughly constant elsewhere.}
  \label{fig:scan-spin-avg}
\end{figure*}

@@

The most severe effect of period on $\sigma(K_{\ell m})$ is in the low-period regime ($P_\omega < 4.3$ hr), but in this case, the most strongly affected parameters are $K_{2m}$, which are generally known better than $K_{3m}$. The effect on the imprecise parameters $K_{3m}$ is small, except for $K_{30}$. It therefore seems as though small-period asteroids are still candidates for observation, although high-period asteroids yield smaller uncertainty.

@@

We assessed the posterior uncertainty generated by the fit process by adjusting various properties of an asteroid on an Earth encounter. The precise thresholds we measure (table \ref{tab:threshold-summary}) show that the observational uncertainty and perigee of the asteroid orbit are the most difficult requirements to meet, with perigee $\lesssim 6$ Earth radii, spin pole uncertainty $\lesssim 1^\circ$, and period uncertainty $\lesssim 10$ ms. However, these thresholds depend on the model used to extract a density distribution, and useful information may be able to be extracted when they are not achieved by using a different model.
